---
title: "Meta-Analysis"
output: html_document
date: '2023-08-10'
---

This R File contains data collated from Data Dryad (Thesis meta-analysis data) as well as a large data file obtained from MacroPopGen database (only microsatellite)
This data contains a total of 16 columns 
1.class 2. threat (non-threatened vs threatened) 3. marker type 4. observed heterozygosity 5. expected heterozygosity 6. mean number of alleles 7. allelic richness 8. fst 9. population stability (given by the IUCN) 10. system (IUCN) 11. movement paper (IUCN) 12. # of threats listed (as counts, IUCN) 13 # of trade use (as counts, IUCN) 14. IUCN RedList status 15. common name 16. scientific name

IUCN STATUS from low concern to high: 
least concern, near threatened, vulnerable, endangered, critically endangered 
IUCN classifies least concern and near threatened as non-threatened species and vulnerable, endangered and critically endangered as threatened species 

This file contains code for:
-check for normality and homogenity of variance within data
-visualization of the data (violin plots)
-one way ANOVA looking differences in each genetic diversity measure amongst IUCN status for each class
-two-sample t test looking at threat level (given by the IUCN) with each genetic diversity measure for each class
-generalized linear regression models looking at each genetic diversity, fst, and population stability as predictor for species threat level

```{r setup, include=FALSE}

#libraries
library(dplyr)
library(ggstance)
library(Hmisc)
library(wesanderson)
library(readxl)
library(ggpubr)
library(car)

#Upload data
meta <- read_excel("~/Desktop/Internship/dryad_macropop_metaanalysis.xlsx")
View(meta)
```

class and iucn redlist rank counts in dataset 
```{r}
#Look at data
head(meta) #total of 16 columns: class, threat, marker, observed, expected, mean alleles, allelic richness, fst, population stability, system, movement pattern, # threats listed, # trade uses listed, IUNC RedList rank
str(meta)
```

```{r}
#obtain counts for each IUCN RedList Ranking across all classes
sum(meta$iucn =="least concern") #5315
sum(meta$iucn =="near threatened") #584
sum(meta$iucn == "vulnerable") #1045
sum(meta$iucn == "endangered") #449
sum(meta$iucn == "critically endangered") #163

#obtain counts for each class (this file has a total of 7,556 entries)
class <- unique(meta$class) 
total <- length(class)
total #six different classes included in this study
sum(meta$class == "Actinopterygii") #2291 total fish data points
sum(meta$class == "Mammalia") #2087 total mammal data points
sum(meta$class == "Aves") #805 total bird data points
sum(meta$class == "Amphibia") #1054 total amphibian data points
sum(meta$class == "Reptilia") #1301 total reptile data points

```

Subset data by class & looking at counts for each IUCN ranking 
```{r}
mammalia <- subset(meta,class=='Mammalia',select=c('observed','expected','mean.alleles', 'allelic.richness', 'fst', 'iucn', 'threat')) #count 2087 
aves <- subset(meta,class=='Aves',select=c('observed','expected','mean.alleles', 'allelic.richness', 'fst', 'iucn', 'threat')) #count 805
amphibia <- subset(meta,class=='Amphibia',select=c('observed','expected','mean.alleles', 'allelic.richness', 'fst', 'iucn', 'threat')) #count 1054
reptilia <- subset(meta,class=='Reptilia',select=c('observed','expected','mean.alleles', 'allelic.richness', 'fst', 'iucn', 'threat')) #count 1301
fish <- subset(meta,class=='Actinopterygii',select=c('observed','expected','mean.alleles', 'allelic.richness', 'fst', 'iucn', 'threat'))

```

```{r}
#obtain counts for each IUCN RedList Ranking for each class
#class mammalia 
sum(mammalia$iucn =="least concern") #1472
sum(mammalia$iucn =="near threatened") #271
sum(mammalia$iucn == "vulnerable") #230
sum(mammalia$iucn == "endangered") #97
sum(mammalia$iucn == "critically endangered") #17

#class aves 
sum(aves$iucn =="least concern") #527
sum(aves$iucn =="near threatened") #109
sum(aves$iucn == "vulnerable") #79
sum(aves$iucn == "endangered") #69
sum(aves$iucn == "critically endangered") #21

#class reptilia
sum(reptilia$iucn =="least concern") #622
sum(reptilia$iucn =="near threatened") #83
sum(reptilia$iucn == "vulnerable") #380
sum(reptilia$iucn == "endangered") #121
sum(reptilia$iucn == "critically endangered") #95
```
#Threat level counts
```{r}
#Actinopterygii
sum(amphibia$threat =="Non Threatened") #890
sum(amphibia$threat =="Threatened") #164

sum(aves$threat =="Non Threatened") #646
sum(aves$threat =="Threatened") #159

sum(fish$threat =="Non Threatened") #1921
sum(fish$threat =="Threatened") #370

sum(reptilia$threat =="Non Threatened") #702
sum(reptilia$threat =="Threatened") #599

sum(mammalia$threat =="Non Threatened") #1744
sum(mammalia$threat =="Threatened") #343

```


visualization of data (all classes combined)
```{r}

meta$iucn <- factor(meta$iucn, #arranges axis by RedList rank
                       c("least concern", "near threatened", "vulnerable", "endangered", "critically endangered"))

#Looking at each genetic diversity measure across all classes 
ob <- ggplot(data = meta, mapping = aes(x = iucn, y = observed))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Observed Heterozygosity") +
  theme_bw()
ob #distribution of observed heterozygosity across IUCN ranking for total data

exp <- ggplot(data = meta, mapping = aes(x = iucn, y = expected))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Expected Heterozygosity") +
  theme_bw()
exp #distribution of expected heterozygosity across IUCN ranking for total data

ma <- ggplot(data = meta, mapping = aes(x = iucn, y = mean.alleles ))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Mean Number of Alleles") +
  theme_bw()
ma #distribution of mean # alleles across IUCN ranking for total data

ar <- ggplot(data = meta, mapping = aes(x = iucn, y = allelic.richness))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Allelic Richness") +
  theme_bw()
ar #distribution of allelic richness across IUCN ranking for total data

fst <- ggplot(data = meta, mapping = aes(x = iucn, y = fst))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Genetic Differentiation (Fst)") +
  theme_bw()
fst #distribution of fst across IUCN ranking for total data

ggarrange(ob, fst, ar, ma + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```


visualization of data (seperating by class: code for Mammals ONLY)
```{r}

meta$iucn <- factor(meta$iucn, #arranges axis by RedList rank
                       c("least concern", "near threatened", "vulnerable", "endangered", "critically endangered"))

#MAMMALIA
mob <- ggplot(data = mammalia, mapping = aes(x = iucn, y = observed))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Observed Heterozygosity") +
  theme_bw()
mob #distribution of observed heterozygosity across IUCN ranking for mammals

mexp <- ggplot(data = mammalia, mapping = aes(x = iucn, y = expected))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Expected Heterozygosity") +
  theme_bw()
mexp #distribution of expected heterozygosity across IUCN ranking for mammals

mma <- ggplot(data = mammalia, mapping = aes(x = iucn, y = mean.alleles ))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Mean Number of Alleles") +
  theme_bw()
mma #distribution of mean # alleles across IUCN ranking for mammals

mar <- ggplot(data = mammalia, mapping = aes(x = iucn, y = allelic.richness))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Allelic Richness") +
  theme_bw()
mar #distribution of allelic richness across IUCN ranking for mamamals

mfst <- ggplot(data = mammalia, mapping = aes(x = iucn, y = fst))+ 
  geom_violin(aes(fill=iucn),show.legend = FALSE)+
  geom_boxplot(width=0.1)+
  xlab("IUCN RedList Ranking") + ylab("Genetic Differentiation (Fst)") +
  theme_bw()
mfst #distribution of fst across IUCN ranking for mammals

ggarrange(mob, mfst, mar, mma + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

Checking for Normality using Shapiro-wilks test
p value < 0.05 indicates that the data is not normal
```{r}
#Observed Heterozygosity
#All failed normality
shapiro.test(mammals$observed)
shapiro.test(aves$observed)
shapiro.test(reptilia$observed)
shapiro.test(amphibia$observed)
shapiro.test(fish$observed)

#Expected Heterozygosity
#All failed normality
shapiro.test(mammals$expected) 
shapiro.test(aves$expected)
shapiro.test(reptilia$expected)
shapiro.test(amphibia$expected)
shapiro.test(fish$expected)

#Mean Alleles
shapiro.test(mammals$mean.alleles) #p=0.099
shapiro.test(aves$mean.alleles)
shapiro.test(reptilia$mean.alleles)
shapiro.test(amphibia$mean.alleles)
shapiro.test(fish$mean.alleles)

#Allelic Richness
#all failed
shapiro.test(mammals$allelic.richness) 
shapiro.test(aves$allelic.richness)
shapiro.test(reptilia$allelic.richness)
shapiro.test(amphibia$allelic.richness)
shapiro.test(fish$allelic.richness)

#FST
#all failed 
shapiro.test(mammals$fst) 
shapiro.test(aves$fst)
shapiro.test(reptilia$fst)
shapiro.test(amphibia$fst)
shapiro.test(fish$fst)
```

#One-Way ANOVA analysis looking at each genetic diversity measures across IUCN rankings
```{r}
#observed heterozygosity
one.way.observed <- aov(observed ~ iucn, data = meta)
summary(one.way.observed) #p-value 2e-16
TukeyHSD(one.way.observed, conf.level = 0.95)

#expected heterozygosity
one.way.expected <- aov(expected ~ iucn, data = meta)
summary(one.way.expected) #p-value 2e-16
TukeyHSD(one.way.expected, conf.level = 0.95)

#mean number of alleles
one.way.ma<- aov(mean.alleles ~ iucn, data = meta)
summary(one.way.ma) #p-value 2e-16
TukeyHSD(one.way.ma, conf.level = 0.95)

#allelic richness
one.way.ar <- aov(allelic.richness ~ iucn, data = meta)
summary(one.way.ar) #p-value 2e-16
TukeyHSD(one.way.ar, conf.level = 0.95)

#fst
one.way.fst <- aov(fst ~ iucn, data = meta)
summary(one.way.fst) #p-value 2e-16
TukeyHSD(one.way.fst, conf.level = 0.95)
```

#Kruskal-Wallis analysis looking at each genetic diversity measure across all classes
(data did not pass normality, need to do nonparametric test)
```{r}
kruskal.test(observed ~ iucn, data = meta)
kruskal.test(expected ~ iucn, data = meta)
kruskal.test(mean.alleles ~ iucn, data = meta)
kruskal.test(allelic.richness ~ iucn, data = meta)
kruskal.test(fst ~ iucn, data = meta)

```

#Kruskal-Wallis analysis looking at each genetic diversity measure for each class 
(data did not pass normality, need to do nonparametric test)
Mammals
```{r}

#observed heterozygosity
one.way.observed <- aov(observed ~ iucn, data = mammals)
summary(one.way.observed) #p-value 0.272 NOT SIG
TukeyHSD(one.way.observed, conf.level = 0.95)

#expected heterozygosity
one.way.expected <- aov(expected ~ iucn, data = mammals)
summary(one.way.expected) #p-value 0.153
TukeyHSD(one.way.expected, conf.level = 0.95)

#mean number of alleles
one.way.ma<- aov(mean.alleles ~ iucn, data = mammals)
summary(one.way.ma) #p-value 2e-16
TukeyHSD(one.way.ma, conf.level = 0.95)

#allelic richness
one.way.ar <- aov(allelic.richness ~ iucn, data = mammals)
summary(one.way.ar) #p-value 2e-16
TukeyHSD(one.way.ar, conf.level = 0.95)

#fst
one.way.fst <- aov(fst ~ iucn, data = mammals)
summary(one.way.fst) #p-value 2e-16
TukeyHSD(one.way.fst, conf.level = 0.95)


```

#### CODE FOR TESTING MODEL WITH NEW INPUT DATA ######
```{r}

pred_test <- predict(model, newdata = data_test, type= "class") 
#Call predict function on mode with new data that has been read in 
pred_test #run to generate prediction 

confusionMatrix(table(pred_test,data_test$target)) #The prediction to compute the confusion matrix and see the accuracy score 

```


