---
title: "Preliminary Stats: Microsatellite Data
output: html_document
date: '2022-08-01'
---

```{r}
library(readxl)
library(ggpubr)
library(car)
mammals <- read_excel("~/Desktop/Thesis/mammals_msat.xlsx")
View(mammals)

mammals$iucn <- factor(mammals$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))


```

#Plots for defense powerpoint
```{r}

mammals$iucn <- factor(mammals$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))

birds$iucn <- factor(birds$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))

fish$iucn <- factor(fish$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))

#Observed Heterozygosity mammals
##ACROSS IUCN RANKING
mo <- ggplot(data=mammals, mapping=aes(x=iucn, y=observed, fill=threat))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()+
  ggtitle("Heterozygosity Across IUCN Rankings in Mammals")+
  theme_bw()+
  theme(legend.position = "right")
mo + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))
#geom_boxplot(outlier.color="black") 

bo <- ggplot(data=birds, mapping=aes(x=iucn, y=observed, fill=threat))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()+
  ggtitle("Heterozygosity Across IUCN Rankings in Birds")+
  theme_bw()
bo + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))
##BY THREAT LEVEL
mot <- ggplot(data=mammals, mapping=aes(x=threat, y=observed, fill=threat))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()+
  ggtitle("Heterozygosity by Threat Level in Mammals")+
  theme_bw()

mot + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))
#geom_boxplot(outlier.color="black") 

bot <- ggplot(data=birds, mapping=aes(x=threat, y=observed, fill=threat))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()+
  ggtitle("Heterozygosity by Threat Level in Birds")+
  theme_bw()

bot + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))



#FST
fstm<- ggplot(data=mammals, mapping=aes(x=iucn, y=fst, fill=threat))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  ggtitle("Fst across IUCN Rankings in Mammals")+
  geom_boxplot()+
  theme_bw()
fstm + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))

fstb<- ggplot(data=birds, mapping=aes(x=iucn, y=fst, fill=threat))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  ggtitle("Fst across IUCN Rankings in Birds")+
  geom_boxplot()+
  theme_bw()
fstb + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))

##BY THREAT
mofst <- ggplot(data=mammals, mapping=aes(x=threat, y=fst, fill=threat))+
  xlab("Threat Level") + ylab("Fst")+ 
  geom_boxplot()+
  ggtitle("Fst by Threat Level in Mammals")+
  theme_bw()

mofst + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))
#geom_boxplot(outlier.color="black") 

fofst <- ggplot(data=fish, mapping=aes(x=threat, y=fst, fill=threat))+
  xlab("Threat Level") + ylab("Fst")+ 
  geom_boxplot()+
  ggtitle("Fst by Threat Level in Fish")+
  theme_bw()

fofst + scale_fill_manual(values = c("#C3D7A4", "#FFDB6D")) + theme(axis.text=element_text(size=10),
        axis.title=element_text(size=14))
```

##################MAMMALS PLOTS##################
```{r plots}
#Observed Heterozygosity 
mo <- ggplot(data=mammals, mapping=aes(x=mammals$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=mammals, mapping=aes(x=threat, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=mammals, mapping=aes(x=mammals$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=mammals, mapping=aes(x=threat, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=mammals, mapping=aes(x=mammals$iucn, y=ar))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=mammals, mapping=aes(x=threat, y=ar))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=mammals, mapping=aes(x=mammals$iucn, y=mean.alleles))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=mammals, mapping=aes(x=threat, y=mean.alleles))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

""" 
#combined plots
ggarrange(mo, fstm, mam, arm + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

#ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
"""
```


#################### MAMMALS DATA Statistical Analysis ########################

#Check Data for Normality Using Shapiro Wilks Test p value < 0.05 indicates that the data is not normal so need to check with log transforming data#
```{r}

shapiro.test(log(mammals$observed)) #did not pass even with log transformed
shapiro.test(log(mammals$fst)) #pass with log transformation
shapiro.test(log(mammals$allelic.richness)) #passed with log tranformation
shapiro.test(mammals$mean.alleles) #Only one that passed normality test

```

##Levenes Test of Homogeneity of variance 
```{r Levene's Test}
#variance across IUCN rank
#leveneTest(observed~iucn, data = mammals) #passed p=0.4785
#leveneTest(log(fst)~iucn, data = mammals) #passed with log transformation p=0.4643
#leveneTest(allelic.richness~iucn, data = mammals) #passed p=0.4159
#leveneTest(mean.alleles~iucn, data = mammals) #passed p-0.5314

#variance across threat level
#leveneTest(observed~threat, data = mammals) #passed p=0.2307
#leveneTest(log(fst)~threat, data = mammals) #passed with log transformation p=0.8686
#leveneTest(allelic.richness~threat, data = mammals) #passed p=0.1945
#leveneTest(mean.alleles~threat, data = mammals) #passed p=0.4242
```

##One Way Anova comparing each genetic diversity measures & fst with IUCN RedList Rankings
```{r Anova}

kruskal.test(observed ~ iucn, data = mammals) #not significantly different p=0.3267

one.way.fst <- aov(log(fst) ~ iucn, data = mammals)
summary(one.way.fst) #no significant p=0.139

one.way.ar <- aov(log(allelic.richness) ~ iucn, data = mammals)
summary(one.way.ar) #not significant p=0.612

one.way.ma <- aov(mean.alleles ~ iucn, data = mammals)
summary(one.way.ma) #not significant p=0.84

#isolated data
#one.way <- aov(observed ~ iucn, data = data_mammals)
#summary(one.way)
#TukeyHSD(one.way, conf.level = 0.95)
#plot(TukeyHSD(one.way, conf.level = 0.99),las=1, col = "red")
```


Variance Test (one sided chi-squared test) and Two-Sample t tests
```{r}

##** remember if p>0.05 we retain the null there is no diff between groups
##log transformed data anti log means e^ (gives you geometric mean)

ob.manwhitney <-wilcox.test(observed ~ threat, data=mammals, na.rm=TRUE, paired=FALSE, exact=FALSE, conf.int=TRUE)
print(ob.manwhitney) #non-parametric man-whitney test on heterozygosity since it was not normally distributed

#res.ftest.fst <- var.test(log(fst)~threat.level, data = mammals)
#res.ftest.fst
#t.test(log(fst), threat.level, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ar <- var.test(log(allelic.richness)~threat.level, data = mammals)
#res.ftest.ar
t.test(log(allelic.richness), threat.level, alternative = "two.sided", var.equal = FALSE)

res.ftest.ma <- var.test(mean.alleles~threat.level, data = mammals)
res.ftest.ma

#t-test for mean alleles since it passed normality 
t.test(mean.alleles, threat.level, alternative = "two.sided", var.equal = FALSE)

```

################### BIRDS DATA #############

```{r}
birds <- read_excel("~/Desktop/Thesis/birds_msat.xlsx")
View(birds)

birds$iucn <- factor(birds$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))
```

```{r plots}
#Observed Heterozygosity 
mo <- ggplot(data=birds, mapping=aes(x=birds$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=birds, mapping=aes(x=threat, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=birds, mapping=aes(x=birds$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=birds, mapping=aes(x=threat, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=birds, mapping=aes(x=birds$iucn, y=allelic.richness))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=birds, mapping=aes(x=threat, y=allelic.richness))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=birds, mapping=aes(x=birds$iucn, y=mean.alleles))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=birds, mapping=aes(x=threat, y=mean.alleles))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

ggarrange(mo, fstm, mam, arm + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

```{r}
shapiro.test(birds$observed) #passed normality p=0.5288
#shapiro.test(log(birds$fst)) #inconclusive
shapiro.test(log(birds$allelic.richness)) #passed with log transformation p=0.0691
shapiro.test(log(birds$mean.alleles)) #failed normality with log
```

##Levene's Test: Homogeneity of Variance All passed
```{r}

#variance across iucn rankings
##all passed 
#leveneTest(observed~iucn, data = birds)
#leveneTest(fst~iucn, data = birds)
#leveneTest(allelic.richness~iucn, data = birds)
#leveneTest(mean.alleles~iucn, data = birds)

#variance across threat level
##all passed 
leveneTest(observed~threat, data = birds) #passed p=0.7068
#leveneTest(fst~threat, data = birds) #did not pass
leveneTest(allelic.richness~threat, data = birds) #passed p=0.06846
leveneTest(mean.alleles~threat, data = birds) #passed p=0.08277

```

One Way Anova: Birds ~ IUCN ranking
```{r}

#one.way.ob <- aov(observed ~ iucn, data = birds) #distributions not identical
#summary(one.way.ob) #significant 0.01639
#TukeyHSD(one.way.ob, conf.level = 0.95) #not significant betwen ranks with ajusted p

#one.way.fst <- aov(log(fst) ~ iucn, data = birds)
#summary(one.way.fst)

#need to run on log transformed data
#one.way.ar <- aov(log(allelic.richness) ~ iucn, data = birds)
#summary(one.way.ar) #not significant p=0.303

#need to run Kruskall-wallis on mean number of alleles because it was not normally distributed
kruskal.test(mean.alleles ~ iucn, data = birds) #p=0.000295
```

```{r, ttest}

##** remember if p>0.05 we retain the null there is no diff between groups
##log transformed data anti log means e^ (gives you geometric mean)

t.test(observed, threat.level, alternative = "two.sided", var.equal = FALSE) #p=2.2-16 

#res.ftest.fst <- var.test(log(fst)~threat.level, data = mmbirds)
#res.ftest.fst
#t.test(log(fst), threat.level, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ar <- var.test(log(allelic.richness)~threat.level, data = mm_mammals)
#res.ftest.ar
t.test(log(allelic.richness), threat.level, alternative = "two.sided", var.equal = FALSE) #p=2.48e-08

#need to run man whitney on mean number of alleles b/c failed normality after transformation
ob.manwhitney <-wilcox.test(mean.alleles ~ threat, data=birds, na.rm=TRUE, paired=FALSE, exact=FALSE, conf.int=TRUE)
print(ob.manwhitney) #significant p=7.141e-06 
```

################## FISH DATA ##############3

```{r}
fish <- read_excel("~/Desktop/Thesis/fish_msat.xlsx")
View(fish)
fish$iucn <- factor(fish$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))
```

```{r plots}
#Observed Heterozygosity 
mo <- ggplot(data=fish, mapping=aes(x=fish$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=fish, mapping=aes(x=threat.level, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=fish, mapping=aes(x=fish$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=fish, mapping=aes(x=threat.level, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=fish, mapping=aes(x=fish$iucn, y=ar))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=fish, mapping=aes(x=threat.level, y=ar))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=fish, mapping=aes(x=fish$iucn, y=m.alleles))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=fish, mapping=aes(x=threat.level, y=m.alleles))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

ggarrange(mo, fstm, mam, arm + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

##Shapiro Wilks Normality Test None of the variables passed normality test
```{r}
shapiro.test(log(fish$observed))#failed
#shapiro.test(log(fish$fst))
shapiro.test(log(fish$allelic.richness)) #passed with log #p=0.9055
shapiro.test(log(fish$mean.alleles)) #passed with log p=0.4303
```

##Levene's Test All of them passed homogeneity of variance test

```{r}
#all passed 
leveneTest(observed~iucn, data = fish)
leveneTest(fst~iucn, data = fish)
leveneTest(log(allelic.richness)~iucn, data = fish)
leveneTest(log(mean.alleles)~iucn, data = fish)
```

##One Way Anova No signficant difference
```{r}
#need to run k.wallis on observed bc failed normality even with log transformation
#kruskal.test(observed ~ iucn, data = fish) #not signicant p=0.2644

#summary(one.way.observed)
#one.way.fst <- aov(fst ~ iucn, data = fish)
#summary(one.way.fst)

#one.way.ar <- aov(log(allelic.richness) ~ iucn, data = fish)
#summary(one.way.ar) #sig 0.0975
#TukeyHSD(one.way.ar, conf.level = 0.95) #adjusted not significant

one.way.ma <- aov(log(mean.alleles) ~ iucn, data = fish)
summary(one.way.ma) #0.0113*
TukeyHSD(one.way.ma, conf.level = 0.95)
```

## Two sample t test
```{r, ttest}

##** remember if p>0.05 we retain the null there is no diff between groups
##log transformed data anti log means e^ (gives you geometric mean)

ob.manwhitney <-wilcox.test(observed ~ threat, data=fish, na.rm=TRUE, paired=FALSE, exact=FALSE, conf.int=TRUE)
print(ob.manwhitney) #not signiciant p=0.1276

#res.ftest.fst <- var.test(fst~threat.level, data = mmfish)
#res.ftest.fst
#t.test(fst, threat.level, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ar <- var.test(log(ar)~threat.level, data = mmfish)
#res.ftest.ar
t.test(log(allelic.richness), threat, data= fish, na.rm=TRUE, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ma <- var.test(log(m.alleles)~threat.level, data = mmfish)
#res.ftest.ma
t.test(log(mean.alleles), threat, data= fish, na.rm=TRUE,alternative = "two.sided", var.equal = FALSE)
```

################## AMPHIBIANS ###################
```{r, read in data}
amp <- read_excel("~/Desktop/Thesis/amphibians_msat.xlsx")
View(amphibians)
```

```{r}
amp <- read_excel("~/Desktop/Pop Gen/mm.amphibians.xlsx")
attach(amp)
amp$iucn <- factor(amp$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))

```
ggplot plots

```{r plots}
#Observed Heterozygosity 
mo <- ggplot(data=amp, mapping=aes(x=amp$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=amp, mapping=aes(x=threat.level, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=amp, mapping=aes(x=amp$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=amp, mapping=aes(x=threat.level, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=amp, mapping=aes(x=amp$iucn, y=ar))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=amp, mapping=aes(x=threat.level, y=ar))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=amp, mapping=aes(x=amp$iucn, y=m.a))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=amp, mapping=aes(x=threat.level, y=m.a))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

ggarrange(mo, fstm, mam, arm + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

##Shapiro Wilks Test: All of them passed test for normality 
```{r}
##all passed normality
shapiro.test(amp$observed)
shapiro.test(amp$fst) 
shapiro.test(amp$allelic.richness)
shapiro.test(amp$mean.alleles)
```

##Levenes Test: All have homogeneity of variance (AR only has one level)
```{r}

# LENGTHS DIFFER COULD NOT RUN HOMOGENEITY OF VARIANCE 

leveneTest(observed~threat, data = amp)
leveneTest(fst~threat, data = amp)
leveneTest(allelic.richness~threat, data = amp)
leveneTest(mean.alleles~threat, data = amp)
```

##Two sample t test##
```{r}
t.test(observed, threat, data=amp, alternative = "two.sided", na.rm=TRUE, var.equal = FALSE)

```

############## REPTILES #######################

```{r}
reptiles <- read_excel("~/Desktop/ConGen/mmreptiles.xlsx")
attach(reptiles)
reptiles$iucn <- factor(reptiles$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))

```

## Reptile plots: Microsatellite Data
```{r plots}
#Observed Heterozygosity 
mo <- ggplot(data=reptiles, mapping=aes(x=reptiles$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=reptiles, mapping=aes(x=threat.level, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=reptiles, mapping=aes(x=reptiles$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=amp, mapping=aes(x=threat.level, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=reptiles, mapping=aes(x=reptiles$iucn, y=ar))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=reptiles, mapping=aes(x=threat.level, y=ar))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=reptiles, mapping=aes(x=reptiles$iucn, y=m.a))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=reptiles, mapping=aes(x=threat.level, y=m.a))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

ggarrange(mo, fstm, mam, arm + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

##Shapiro Wilks Test: All of them passed test for normality 
```{r}
shapiro.test(reptiles$observed)
shapiro.test(reptiles$fst) 
shapiro.test(reptiles$ar)
shapiro.test(reptiles$m.a)
```

##Levenes Test: All have homogeneity of variance (AR only has one level)
```{r}
leveneTest(observed~iucn, data = reptiles)
leveneTest(fst~iucn, data = reptiles)
leveneTest(ar~iucn, data = reptiles)
leveneTest(m.a~iucn, data = reptiles)
```

################# INSECTS ###############
```{r}
insects <- read_excel("~/Desktop/ConGen/mminsects.xlsx")
insects$iucn <- factor(insects$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))
```

##Shapiro Wilks Test: all passed (ar: transformed)
```{r}
shapiro.test(insects$observed)
shapiro.test(log(insects$fst) 
shapiro.test(insects$ar)
shapiro.test(insects$mean.alleles)
```

##Levenes Test: All passed
```{r}
leveneTest(observed~iucn, data = insects)
leveneTest(log(fst)~iucn, data = insects)
leveneTest(ar~iucn, data = insects)
leveneTest(mean.alleles~iucn, data = insects)

```

```{r}
one.way.observed <- aov(observed ~ iucn, data = insects)
summary(one.way.observed)
one.way.fst <- aov(log(fst) ~ iucn, data = insects)
summary(one.way.fst)
one.way.ar <- aov(ar ~ iucn, data = insects)
summary(one.way.ar)
one.way.ma <- aov(mean.alleles ~ iucn, data = insects)
summary(one.way.ma)
TukeyHSD(one.way.ma, conf.level = 0.95)
```

## Two sample t test

```{r, ttest}

##** remember if p>0.05 we retain the null there is no diff between groups
##log transformed data anti log means e^ (gives you geometric mean)
#library("ggpubr")
#install.packages("ggpubr")

#res.ftest.fst <- var.test(fst~threat.level, data = mmfish)
#res.ftest.fst
#t.test(fst, threat.level, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ar <- var.test(log(ar)~threat.level, data = mmfish)
#res.ftest.ar
t.test(ar, threat.level, alternative = "two.sided", var.equal = FALSE)

#res.ftest.ma <- var.test(log(m.alleles)~threat.level, data = mmfish)
#res.ftest.ma
t.test(mean.alleles, threat.level, alternative = "two.sided", var.equal = FALSE)
```

##PLOTS
```{r}
insects <- read_excel("~/Desktop/Thesis/insects_msat.xlsx")
insects$iucn <- factor(insects$iucn,
                       c("Least Concern", "Near Threatened", "Vulnerable", "Endangered", "Critically Endangered"))


#Observed Heterozygosity 
mo <- ggplot(data=insects, mapping=aes(x=insects$iucn, y=observed))+
  xlab("ICUN Ranking") + ylab("Observed Heterozygosity")+ 
  geom_boxplot()
mo 

ntmo <- ggplot(data=insects, mapping=aes(x=threat.level, y=observed))+
  xlab("Threat Level") + ylab("Observed Heterozygosity")+
  geom_boxplot()
ntmo

#FST
fstm<- ggplot(data=insects, mapping=aes(x=insects$iucn, y=fst))+
  xlab("ICUN Ranking") + ylab("Fst")+ 
  geom_boxplot()
fstm
ntfstm <- ggplot(data=insects, mapping=aes(x=threat.level, y=fst))+
  xlab("Threat Level") + ylab("Fst") +
  geom_boxplot()
ntfstm

#Allelic Richness
arm<- ggplot(data=insects, mapping=aes(x=insects$iucn, y=ar))+
  xlab("ICUN Ranking") + ylab("Allelic Richness")+ 
  geom_boxplot()
arm
ntarm<- ggplot(data=insects, mapping=aes(x=threat.level, y=ar))+
  xlab("Threat Level") + ylab("Allelic Richness")+
  geom_boxplot()
ntarm

#Mean Number of Alleles
mam <- ggplot(data=insects, mapping=aes(x=insects$iucn, y=mean.alleles))+
  xlab("ICUN Ranking") + ylab("Mean Number of Alleles")+ 
  geom_boxplot()
mam
ntmam<- ggplot(data=insects, mapping=aes(x=threat.level, y=mean.alleles))+
  xlab("Threat Level") + ylab("Mean Number of Alleles")+
  geom_boxplot()
ntmam

ggarrange(mo, fstm, mam, arm + rremove("x.text"),
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)

ggarrange(ntmo, ntfstm, ntarm, ntmam + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```



