---
title: "Phylogeny inference analysis: Microsatellite Data"
output: html_document
date: '2023-01-24'
---

```{r setup, include=FALSE}

setwd("~/Desktop/ConGen")

library(readxl)
library(scales)
library(ggplot2)
library(see)
library(tidyverse)
library(nlme)
library(ape)
library(phytools)
library(caper)
library(geiger)
library(dplyr)

```

#Microsatellite Data#
```{r}

animals <- read_excel("mammals_msat.xlsx", sheet = 1, na = c("", "NA"))
animals$sci.name = paste(animals$Genus, animals$Species, sep=" ") #sci name

## Set up data: Mammals dataset ##
animals$Phylum  = as.factor(as.character(animals$Phylum)) #as.factor converts to signify levels 
animals$Class   = as.factor(as.character(animals$Class)) #as.character converts object with letters
animals$Order   = as.factor(as.character(animals$Order))
animals$Family  = as.factor(as.character(animals$Family))
animals$Genus   = as.factor(as.character(animals$Genus))
animals$Species = as.factor(as.character(animals$Species))
animals$iucn[animals$iucn==""] = NA
animals$iucn    = as.factor(as.character(animals$iucn))

```

##Set up phylogenetic Tree
```{r}
#as phylo.formula2 builds tree from a set of nested taxonomic variables
as.phylo.formula2 = function (x, data = parent.frame(), collapse = TRUE){ 
  err <- "Formula must be of the kind \"~A1/A2/.../An\"."
  if (length(x) != 2) 
    stop(err)
  if (x[[1]] != "~") 
    stop(err)
  f <- x[[2]]
  taxo <- list()
  while (length(f) == 3) {
    if (f[[1]] != "/") 
      stop(err)
    if (!is.factor(data[[deparse(f[[3]])]])) 
      stop(paste("Variable", deparse(f[[3]]), "must be a factor."))
    taxo[[deparse(f[[2]])]] <- data[[deparse(f[[3]])]]
    if (length(f) > 1) 
      f <- f[[2]]
  }
  if (!is.factor(data[[deparse(f)]])) 
    stop(paste("Variable", deparse(f), "must be a factor."))
  taxo[[deparse(f)]] <- data[[deparse(f)]]
  taxo.data <- as.data.frame(taxo)
  leaves.names <- as.character(taxo.data[, 1])
  taxo.data[, 1] <- 1:nrow(taxo.data)
  f.rec <- function(subtaxo) {
    u <- ncol(subtaxo)
    levels <- unique(subtaxo[, u])
    if (u == 1) {
      if (length(levels) != nrow(subtaxo)) 
        warning("Error, leaves names are not unique.")
      return(as.character(subtaxo[, 1]))
    }
    t <- character(length(levels))
    for (l in 1:length(levels)) {
      x <- f.rec(subtaxo[subtaxo[, u] == levels[l], ][1:(u - 1)])
      t[l] <- paste("(", paste(x, collapse = ","), ")", sep = "")
    }
    return(t)
  }
  string <- paste("(", paste(f.rec(taxo.data), collapse = ","),");", sep = "")
  
  #read.newick reads a newick style tree into file 
  phy <- read.newick(text = string) ## so that singles will be read without error
  phy$edge.length <- rep(1,nrow(phy$edge))
  phy <- collapse.singles(phy)
  phy$tip.label <- leaves.names[as.numeric(phy$tip.label)]
  return(phy)
}

library(phytools)
data(animals)
t1<-as.phylo.formula(~Phylum/Class/Order/Family/Genus/Species, 
    data=animals)
plotTree(t1,fsize=0.7,ftype="i")
```

```{r}
##observed heterozyogsity## 
#phylo uses phylogeny inference using least squares 
animalsdf = animals[!is.na(animals$iucn) & !is.na(animals$ar),]
animalstreeT <- as.phylo.formula(~Phylum/Class/Order/Family, data = animals, collapse = TRUE) 
animalsTree <- multi2di(animalstreeT)
#we need to look at residuals of distribution of observed heterozygosity from taxonomy to see correlation from shared ancestry
#are errors correlated? 
outpgls = gls(ar~iucn-1, correlation = corAR1(form = ~1 | sci.name), data = animalsdf, method = "ML")
#method ML -> log liklihood is maximized (data not normally dist)
summary(outpgls) #correlation of linear model using least squares 
#standardized residual errors

#residual std error/root n to give stdev 
#squared of stdev to get variance

```

animalsdf = animals[!is.na(animals$iucn) & !is.na(animals$observed),]
animalstreeT <- as.phylo.formula(~Phylum/Class/Order/Family, data = animals, collapse = TRUE) 
animalsTree <- multi2di(animalstreeT)
outpgls = gls(observed~iucn-1, correlation = corAR1(form = ~1 | sci.name), data = animalsdf, method = "ML")
summary(outpgls)

animalsdf = animals[!is.na(animals$iucn) & !is.na(animals$fst),]
animalstreeT <- as.phylo.formula(~Phylum/Class/Order/Family, data = animals, collapse = TRUE) 
animalsTree <- multi2di(animalstreeT)
outpgls = gls(fst~iucn-1, correlation = corAR1(form = ~1 | sci.name), data = animalsdf, method = "ML")
summary(outpgls)


```

